Motto -- Attributed to John Chambers, ACM award winner, father of S and R:

In R,

- Everything that **exists** is an object;
- Everything that **happens** is a function call

Consequently, we want to

1. understand R objects
2. see *what* functions are called when _"we type simple things"_

## **Replacement functions :**

```{r}
v <- 1:5
names(v) <- LETTERS[1:5]
v
v <- 1:5
names(v) <- c("Hans", "Heiri")
v
names(v) # did construct character NA's
```

Hmm, ...,  what functions where called in the above lines?

```{r}
`names<-`
get("names<-") # here equivalent to the line above, using   ` ... `
identical(`names<-`, get("names<-"))
```

### Excursion: Look at the functions in `base` etc:
```{r}
search()
?search
?ls
ls()
ls(2)  ## (and mention the Rstudio GUI to it)
```

>>> Open  (`packages.Rnw` or rather its small excerpt) `packages-search.Rmd`

## Replacement functions -2-  `someT<-`

Now, we know this:
We can define quite funny functions ourself ..
```{r}
`[<-`  <- function(){}
find("[<-") # in two places ..
get("[<-",1)
get("[<-","package:base")

x <- 1:7
try(
x[1] <- 10
)# --> an error .. oops !!
```
we broke  R .... well, *our* R session only.
How to fix?




```{r}
rm("[<-")
x[1] <- 10 ; x ## aaahh... life is well again

`[<-.factor` ## the "factor" method of the `[<-` generic function
 ## This is the "factor" method of the `[` generic function
methods("[<-")
```

## **Note**: The three parts of all (regular) R functions:

An R function is made of three parts
1. `formals()`    : The argument list
2.    `body()`    : The function body
3. `environment()`: The environment in which the function has been defined

```{r}
str(abline) # the function "header" *printed*
args(abline)  # the "[arg]ument[s]" (but ..)
str(args(abline)) # {as empty function}
f <- abline
body(f) <- NULL
f

## The "real" argument list as R object:
str ( formals(abline) ) # "Dotted pair list"

?body
?formals ##  there is a replacement function
def.u  <- formals(abline)$untf
def.u
deff <- formals(formals)[["fun"]]
deff
class(deff)
```

### `missing()` allows to check if an argument has been specified  in the actual call:
```{r}
tt  <- function(x, ab=TRUE) {
        cat("missing(x):", missing(x),
            "miss.(ab):", missing(ab),"\n")
        formals()
}
tt(1)
tt(1, TRUE)
tt() ## the formals arguments of tt()
```

### Formal (and "actual") Arguments

```{r}
sin     # formal argument  'x'
sin(pi) # actual argument 'pi' --- is matched to the formal:
```
... something like `x <- pi` happens implicitly when the function is
called: The formal argument is replaced by the actual.

Beware of  "Lazy Evaluation" (replacement only happens as needed)

sin is  *Primitive* function -- these are made to be very fast in R
by ``diving into C-level'' of R immediately.
==> The primitive functions are **the exceptions**: all three parts are empty
```{r}
formals(sin) ## NULL, ditto for the body
body   (sin)
environment(sin)
```
