---
title: Language Objects in R
author: Martin Maechler
date: "Aug. 28, 2015  (last change: Mar. 2, 2021; rendered at `r Sys.Date()`)" # format(Sys.time(), "%H:%M:%OS3")
output:
  html_document:
    highlight: tango
    keep_md: yes
    theme: journal
---
# R's building blocks

In one way, since in R,

>    *Everything that happens is a function call*   (John Chambers, part 2),

`call`s are the basic language objects in R.

```{r sin_x}
cc <- quote( sin(x) )
cc
class(cc)
```

Looking closely, `call`s are made from `symbol`s, here, `sin(x)` is made
from the two symbols `sin` and `x`:
```{r calls2}
length(cc)
c1 <- cc[[1]]; c1
c2 <- cc[[2]]; c2

typeof(c1)
typeof(c2)
   str(c2)
class(c1)
```

# Get to the substance - R functions of object inspection

Since in R,

>    *Everything that exists is an object*     (John Chambers, part 1),

we want to find out what "kind" of an object we have.
The most useful "user-level" function to get an idea about an object is `str()` for displaying the **str**ucture of an R object...  It is used very often interactively, to get both a bit of an overview and some details about object's structure.
`str()` is not meant to be used as tool to be used in other functions, so from a programmeR's view,
the standard functions for object inspection to use are `class()` which is "high-level" and `typeof()` which is low-level.
For historical reasons (back compatibility with **S** / **S+**), another low-level inspector function is `storage.mode()`  and another higher-level one is `mode()`.

Here, we use them on a somewhat representative collection of R objects (all conveniently put in a `list`)
to see how these R objects, are categorized by the different "inspection" functions.

<!-- FIXME: start with expression(..., .., .., ) instead of list(.., ) ==> -->
<!-- even nicer table-->
```{r l.ex}
l.ex <- list(one = 1, s1 = 1:2, pi = pi, I = 1i, `NA` = NA, let = c("A", "a", "b", ":::"),
             fn = mean, fn2 = c, fnSpec = `function`,
             n = as.name("Mä"), n2 = quote(x), nb = quote(`{`),
             ex = expression(1+1), cl = call("round",10), cl2 = quote(sin(x)),
             formula = y ~ x+t, formals = formals(lm), arg1 = formals(lm)[[1]],
             NULL = NULL)
str(l.ex, vec.len = 16, max.level = 1)
```

```{r show.them}
myShow <- function(x, max.length= 1000L) {
    r <- tryCatch(format(x), error=function(e)e)
    r <- if(inherits(r, "error"))
             tryCatch(as.character(x), error=function(e)e)
         else paste(r, collapse = " ")
    r <- if(inherits(r, "error"))
             tryCatch(capture.output(x), error=function(e)e)
         else paste(r, collapse = " ")
    substr(r, 1L, max.length)
}

cbind(show    = sapply(l.ex, myShow, max.length = 16),
      typeof  = sapply(l.ex, typeof),
      st.mode = sapply(l.ex, storage.mode),
      mode    = sapply(l.ex, mode),
      class   = sapply(l.ex, class)) -> tab.ex                                                                          #=====
```

Now instead of printing this directly - which is readable and nice already,
let's be "advanced" and use a `knitr` package feature to produce "raw markdown" tables:
```{r table-show, results='asis'}
knitr::kable(tab.ex, format="pandoc")
```
```{r session-info}
print(sessionInfo(), locale=FALSE)
```

<!-- Note the default of the  `backtick` argument of function `deparse()`.
 It currently is `mode(expr) %in% c("call", "expression", "(", "function")`.
 I had tried to change this (and add "name" to the strings) in Sep.2017,
 but that broke a couple of (not unimportant) CRAN packages.
I found that this unfortunately meant the `symbol`s (aka `name`s) would not be correctly deparsed by default.  E.g., the above `nb` would not by default use `backtick = TRUE` but that would have been quite desirable.
 -->

_Note_: The source of this document is MM's `r knitr::current_input(dir=TRUE)`
